Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
remove bullet
Related work

Maximize the ELBO or video reconstruction error to learn a world model can mislead the model to focus on learning task-irrelevant information, which distorts the optimization landscape. When the task is complex, planning with reconstruction-based world models is unreliable \cite{lutter2021learning} and struggles to outperform model-free methods \cite{kostrikov2020image, yarats2021mastering}. Planning can be done through models that learn to predicting rewards and values for multi-step rollouts \cite {tamar2016value, silver2017predictron, oh2017value, schrittwieser2020mastering}. However, these methods do not utilize the observations for model learning at all, which can make them sample in-efficient.

How to use observations efficiently in model-based RL without resorting to reconstruction or direct prediction of future frames? \citet{ye2021mastering} introduce a self-supervised consistency loss on the learned latents. Instead of relying on a reconstruction-based loss function, \citet{schwarzer2020data} propose a cosine similarity loss between predicted future latents and the true future latents, and performing Q-learning in the learned latent space. \citet{ma2021contrastive} modify the classical world model learning objective by replacing the observation reconstruction loss with a contrastive learning term, which maximizes the mutual information between latent states and observations and helps combat the model focusing on learning task-irrelevant information. \citet{kipf2019contrastive} use contrastive learning to learn a structured world model. The contrastive objectives for model-based RL significantly improve sample efficiency, but they are not only limited to model-based reinforcement learning. \citet{laskinCURLContrastiveUnsupervised2020} propose replacing the transition function with data augmentations to get a contrastive target, and performing model-free learning using the representation produced by the encoder trained with the contrastive target.

@@

\textbf{\our does not suffer from rank collapse} We analyze the impact of the latent-space normalization on the rank of the latent representations outputted by the learned encoder. We follow the methodology of \citet{ni2024bridging} for estimating the matrix rank. We plot the ranks in \cref{fig:rank-comparison-z1}. The latent state dimensions are 256, 128, 256, and 128 for Acrobot Swingup, Hopper Stand, Humanoid Run and Cheetah Run, respectively, and the plots show that the matrix rank stays close to the latent state dimension when our FSQ normalization scheme is applied. Without normalization, a dimensional collapse occurs, which can have significant harmful effects as the representational power of the latent state diminishes \cite{jingUnderstandingDimensionalCollapse2021}.

@@

\textbf{Limitations and future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research in multi-task and lifelong learning.
Given that our method obtains an implicit codebook an obvious future direction is exploring if we can leverage the
discretized latent-space. So far, we have only applied our representation learning loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an interesting direction. Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample efficient. Exploring whether explicitly quantifying the model uncertainty by \eg\ using larger ensembles would help to adapt our method to a real-world robotics setting is also a promising direction for future work.

% Arno recommends to have this here as a last thing (leaves a positive feeling)
A reference implementation of the methods will be made available on GitHub upon acceptance of the paper (see supplementary material for now).

>>>>

Maximizing the ELBO or video reconstruction error to learn a world model can mislead the model to focus on learning task-irrelevant information, which distorts the optimization landscape \cite{zhang2018natural,zintgraf2021varibad}. When the task is complex, planning with reconstruction-based world models is unreliable \cite{lutter2021learning} and struggles to outperform model-free methods \cite{kostrikov2020image, yarats2021mastering}. The problems are aggravated when the environment is partially observable \cite{morad2023popgym}. Planning can be done through models that learn to predicting rewards and values for multi-step rollouts \cite {tamar2016value, silver2017predictron, oh2017value, schrittwieser2020mastering}. However, these methods do not utilize the observations for model learning at all, which can make them sample in-efficient.

How to use high-dimensional and noisy observations efficiently in model-based RL without resorting to reconstruction or direct prediction of future frames? A common solution has been to attach an auxiliary loss to the RL objective and perform representation learning \cite{tomar2021learning}. A common approach for learning suitable representations is the use of self-predictive abstractions, where the model is trained to predict future latent states through an auxiliary loss \cite{subramanian2022approximate, ni2024bridging}. \citet{ye2021mastering} introduce a self-supervised consistency loss on the learned latents. Instead of relying on a reconstruction-based loss function, \citet{schwarzer2020data} propose a cosine similarity loss between predicted future latents and the true future latents, and performing Q-learning in the learned latent space. \citet{ma2021contrastive} modify the classical world model learning objective by replacing the observation reconstruction loss with a contrastive learning term, which maximizes the mutual information between latent states and observations and helps combat the model focusing on learning task-irrelevant information. \citet{kipf2019contrastive} use contrastive learning to learn a structured world model. \citet{ghugare2022simplifying} argue that aligning the auxiliary objectives and RL objective can be difficult, and derive a bound directly on the overall RL objective, while is compatible with self-consistency. The contrastive objectives for model-based RL significantly improve sample efficiency, but they are not only limited to model-based reinforcement learning. \citet{laskinCURLContrastiveUnsupervised2020} propose replacing the transition function with data augmentations to get a contrastive target, and performing model-free learning using the representation produced by the encoder trained with the contrastive target.

@@

\textbf{\our does not suffer from rank collapse} We analyze the impact of the latent-space normalization on the rank of the latent representations outputted by the learned encoder. We follow the methodology of \citet{ni2024bridging} for estimating the matrix rank. We plot the ranks in \cref{fig:rank-comparison-z1}. The latent state dimensions are 256, 128, 256, and 128 for Acrobot Swingup, Hopper Stand, Humanoid Run and Cheetah Run, respectively, and the plots show that the matrix rank stays close to the latent state dimension when our FSQ normalization scheme is applied. Without normalization, a dimensional collapse occurs, which can have significant harmful effects as the representational power of the latent state diminishes \cite{jingUnderstandingDimensionalCollapse2021}. Note that in contrast to \citet{ni2024bridging}, we observe that using a detached momentum encoder as the target encoder is insufficient to prevent a rank collapse in practice. We hypothesize that this is due to trivial representations being enough to minimize the prediction error, when a reward term is not present in the loss \cite{tang2023understanding}.

@@

\textbf{Limitations and Future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research in multi-task and lifelong learning.
Given that our method obtains an implicit codebook an obvious future direction is exploring if we can leverage the
discretized latent-space. So far, we have only applied our representation learning loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an interesting direction. Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample efficient. Exploring whether explicitly quantifying the model uncertainty by \eg\ using larger ensembles would help to adapt our method to a real-world robotics setting is also a promising direction for future work.
